---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CollegeMetricsR

<!-- badges: start -->

<!-- badges: end -->

**CollegeMetricsR** is an R package designed to analyze and visualize the relationship between student lifestyle habits and academic performance. The package focuses on key factors such as study time, sleep, attendance, and extracurricular involvement, and provides tools to explore how these variables relate to outcomes like exam scores and GPA.

College students often balance multiple responsibilities, and it can be difficult to understand how these factors interact. This package simplifies that process by offering intuitive functions for summarization, visualization, correlation analysis, and performance evaluation.

While there are datasets on student life, there are a limited number of beginner-friendly tools that:

-   Structure student lifestyle data

-   Analyze relationships between variables

-   Help students predict & improve their performance

**CollegeMetricsR** aims to fill in this gap by providing a comprehensive set of functions that will help analyze student lifestyle factors.

## Installation

You can install the development version of CollegeMetricsR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("anikavb/CollegeMetricsR")
```

```{r example}
library(CollegeMetricsR)
```

All required dependencies (such as `ggplot2`) are automatically installed when installing the package.

## Data Set

<https://www.kaggle.com/datasets/lainguyn123/student-performance-factors/code>

The package utilizes a real-world dataset from Kaggle containing student performance factors, including study habits, sleep patterns, and external influences.

This dataset includes both numerical and categorical variables, such as hours studied, sleep

duration, parental involvement, and exam scores. Within the package, the data will be cleaned and transformed as needed to facilitate analysis of relationships between lifestyle factors and academic performance.

## Functionality

This package enables users to:

-   Explore student performance data

-   Examine relationships between lifestyle variables and academic outcomes

-   Summarize key patterns in student data

-   Visualize trends related to student performance across metrics

-   Estimate academic performance based on personal habits

The package incorporates the following functions:

### Data Preparation

-   `load_student_data()` Loads and cleans the student dataset included in the package.

-   `scale_exam_score()` Converts exam scores into a GPA-like scale for easier interpretation.

### Summary & Exploration

-   `lifestyle_summary()` Provides summary statistics (mean, median, sd, etc.) for key student variables.

-   `sleep_summary()` Summarizes sleep patterns across students.

-   `grade_analysis()` Explores GPA/grade distribution across students.

-   `student_correlation()` Calculates the correlation between any two numeric variables and provides a simple interpretation.

### Visualization

-   `plot_relationship()` Creates scatterplots between any two variables with a regression trend line.

-   `plot_study_vs_gpa()` Visualizes the relationship between study time and GPA.

-   `plot_sleep_vs_gpa()` Visualizes how sleep patterns relate to GPA.

### Performance Insights

-   `performance_score()` Generates a composite performance score based on GPA, study time, sleep, attendance, and extracurricular involvement. Also provides suggestions for improvement.

-   `predict_gpa()` Predicts a student’s GPA based on lifestyle inputs using a linear model.

-   `top_performers()` Identifies the top students based on a selected metric (e.g., exam score or GPA)

## Example Workflow

```{r}

# Load and prepare data
df <- load_student_data()
df <- scale_exam_score(df)

# Summarize key variables
lifestyle_summary(df)

# Visualize relationships
plot_relationship(df, "Hours_Studied", "gpa")

# Analyze correlation
student_correlation(df, "Sleep_Hours", "Hours_Studied")

# Predict GPA
predict_gpa(
  hours_studied = 20,
  sleep_hours = 7,
  attendance = 85,
  previous_scores = 75,
  motivation_level = "Medium"
)

# Compute performance score
performance_score(
  gpa = 3.5,
  hours_studied = 20,
  attendance = 90,
  sleep_hours = 7,
  extracurricular = "Yes"
)

# Identify top students
top_performers(df, n = 5, metric = "gpa")

```

## Use Cases

This package is useful for:

-   Students analyzing their own academic habits and searching for ways to improve.

-   Researchers exploring relationships between lifestyle and performance.

-   Beginners learning data analysis in R Demonstrating simple statistical and visualization workflows.
