---
title: "Getting Started with CollegeMetricsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CollegeMetricsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CollegeMetricsR)
```

## Overview

CollegeMetricsR is an R package designed to help users explore, analyze, and visualize the relationship between student lifestyle habits and academic performance. It is built around a real dataset of 6,607 students from Kaggle, containing variables such as sleep hours, study time, attendance, motivation level, and exam scores.

The package is designed to be beginner-friendly. You do not need to write any data wrangling code yourself. The functions follow a natural pipeline from loading data all the way through to predicting scores.

This vignette walks through a complete analysis using the package's dataset and all of its core functions, in the order you would naturally use them: load and inspect, summarize, explore relationships, engineer features, identify standouts, and finally predict.

------------------------------------------------------------------------

## Step 1: Load the Data

The dataset is built into the package as `student_data`. You can load a cleaned version of it using `load_student_data()`, which automatically removes any rows with missing values.

```{r load}
df <- load_student_data()
nrow(df)
head(df[, c("Hours_Studied", "Sleep_Hours", "Attendance",
            "Motivation_Level", "Exam_Score")])
```

The dataset contains 20 variables covering lifestyle habits, personal background, and academic outcomes. For most functions in this package the key columns are `Hours_Studied`, `Sleep_Hours`, `Attendance`, `Previous_Scores`, `Motivation_Level`, and `Exam_Score`.

------------------------------------------------------------------------

## Step 2: Convert Exam Scores to GPA

Raw exam scores in this dataset run from 0 to 100. To make them easier to interpret, `scale_exam_score()` converts them to a GPA scale (0.0 to 4.0) using the formula `(Exam_Score / 100) * 4`.

```{r gpa}
df <- scale_exam_score(df)
head(df[, c("Exam_Score", "gpa")])
```

This adds a new `gpa` column to the data frame which is used by several other functions downstream.

------------------------------------------------------------------------

## Step 3: Summarize Lifestyle Variables

Before exploring relationships, it helps to understand the distribution of each variable. `lifestyle_summary()` computes mean, median, standard deviation, min, max, and a missing-value count for all numeric columns at once.

```{r lifestyle-summary}
lifestyle_summary(df)
```

You can also pass a `vars` argument to focus on specific columns:

```{r lifestyle-summary-vars}
lifestyle_summary(df, vars = c("Hours_Studied", "Sleep_Hours", "Exam_Score", "Attendance"))
```

### Additional Quick Summaries

The package provides convenient functions for two commonly examined distributions in student data.

`grade_analysis()` gives a quick overview of the GPA distribution across all students.

```{r grades}
grade_analysis(df)
```

This gives you a quick sense of where most students fall on the GPA scale and how much variation there is.

`sleep_summary()` returns the mean, median, and standard deviation of student sleep hours across the whole dataset.

```{r sleep-summary}
sleep_summary(df)
```

------------------------------------------------------------------------

## Step 4: Explore correlations between variables

`variable_correlation()` calculates the correlation between any two numeric variables in the dataset. It returns the correlation coefficient along with a clear, plain-language interpretation of the strength and direction of the relationship.

```{r correlation-study-gpa}
student_correlation(df, "Hours_Studied", "gpa")
```

This enables users to see at a glance which lifestyle factors are most strongly associated with exam outcomes.

For specificity, `sleep_study_relationship()` computes the correlation between sleep hours and hours studied, and gives a plain-language interpretation.

```{r sleep-study}
sleep_study_relationship(df)
```

A correlation close to 0 here would suggest that how much a student sleeps has little relationship with how much they study — students who study more are not necessarily sleeping less.

------------------------------------------------------------------------

## Step 5: Visualize Relationships

To visually interpret the correlations above, `plot_relationship()` lets you plot any two numeric variables against each other without writing any ggplot2 code yourself. You simply specify the variables you want to compare.

```{r plot-relationship}
plot_relationship(df, "Hours_Studied", "gpa")
```

### Additional Visualization Functions

In addition to the flexible plotting function, the package also includes predefined plots for common relationships.

**Sleep vs GPA**

```{r plot-sleep}
plot_sleep_vs_gpa(df)
```

This visualization highlights how sleep patterns may influence academic outcomes, complementing the numerical summaries from earlier steps.

**Study Time vs GPA**

```{r plot-study}
plot_study_vs_gpa(df)
```

This plot focuses specifically on the relationship between study time and GPA, making it easy to interpret one of the most important factors in academic performance.

Together, these visualization functions allow users to both explore custom relationships and quickly analyze key patterns built into the package.

------------------------------------------------------------------------

## Step 6: Predict GPA

`predict_gpa()` trains a linear regression model on the full dataset and predicts GPA based on a new student's personal inputs. You just provide your own values directly, no need to build a data frame.

```{r predict}
predict_gpa(hours_studied    = 20,
            sleep_hours      = 7,
            attendance       = 85,
            previous_scores  = 75,
            motivation_level = "Medium")
```

You can try different inputs to see how the prediction changes:

```{r predict-compare}
# High performing student
predict_gpa(hours_studied    = 35,
            sleep_hours      = 9,
            attendance       = 98,
            previous_scores  = 95,
            motivation_level = "High")

# Struggling student
predict_gpa(hours_studied    = 5,
            sleep_hours      = 4,
            attendance       = 60,
            previous_scores  = 45,
            motivation_level = "Low")
```

The model uses hours studied, sleep hours, attendance, previous scores, and motivation level as predictors. It is a simple linear model - the intention is to be able interpret the GPA rather than high predictive accuracy.

------------------------------------------------------------------------

## Step 7: Calculate Your Performance Score

`performance_score()` combines key lifestyle variables into a single score between 0 and 100. It also prints a performance message and personalised suggestions based on your inputs.

Weights used:

-   GPA: 30%
-   Hours Studied: 25%
-   Attendance: 15%
-   Sleep Hours: 15%
-   Extracurriculars: 15%

```{r performance-average}
# Average student
performance_score(
  gpa             = 2.7,
  hours_studied   = 15,
  attendance      = 75,
  sleep_hours     = 7,
  extracurricular = "No"
)
```

```{r performance-high}
# High achieving student
performance_score(
  gpa             = 3.8,
  hours_studied   = 35,
  attendance      = 95,
  sleep_hours     = 8,
  extracurricular = "Yes"
)
```

```{r performance-low}
# Struggling student
performance_score(
  gpa             = 1.5,
  hours_studied   = 5,
  attendance      = 60,
  sleep_hours     = 5,
  extracurricular = "No"
)
```

The composite score gives a more rounded picture of each student than a single exam result alone. The function not only returns a score but also prints tailored suggestions based on average metrics, so users know exactly which habits to improve.

------------------------------------------------------------------------

## Step 8: Identify top and bottom performers

`top_performers()` returns the top `n` students ranked by any numeric column, with a `rank` column prepended. This allows users to analyze the overall profile of the top students for a given metric.

By default, it ranks students by `Exam_Score`.

```{r top-exam}
top_performers(df, n = 5)
```

Say you wanted to explore by `Hours_Studied`:

```{r top-study}
top_performers(df, n = 5, metric = "Hours_Studied")
```

This can help users explore what characteristics are common among high-performing students.

## Conclusion

**CollegeMetricsR** provides a simple workflow for analyzing student performance data. Users can load and prepare data, summarize lifestyle variables, visualize relationships, calculate correlations, predict GPA, compute a performance score, and identify top performers.\
\
The package is designed to be beginner-friendly and interpretable. Future improvements could include more advanced modeling, additional datasets, and more personalized recommendations.
