---
title: "Getting Started with CollegeMetricsR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CollegeMetricsR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CollegeMetricsR)
```

## Introduction

CollegeMetricsR is an R package designed to help users explore, analyze,
and visualize the relationship between student lifestyle habits and academic
performance. It is built around a real dataset of 6,607 students from Kaggle,
containing variables such as sleep hours, study time, attendance, motivation
level, and exam scores.

The package is designed to be beginner-friendly.You do not need to write
any data wrangling code yourself. The functions follow a natural pipeline
from loading data all the way through to predicting GPA.

This vignette walks through that full pipeline step by step.

---

## Step 1: Load the Data

The dataset is built into the package as `student_data`. You can load a
cleaned version of it using `load_student_data()`, which automatically
removes any rows with missing values.

```{r load}
df <- load_student_data()
nrow(df)
head(df[, c("Hours_Studied", "Sleep_Hours", "Attendance",
            "Motivation_Level", "Exam_Score")])
```

The dataset contains 20 variables covering lifestyle habits, personal
background, and academic outcomes. For most functions in this package
the key columns are `Hours_Studied`, `Sleep_Hours`, `Attendance`,
`Previous_Scores`, `Motivation_Level`, and `Exam_Score`.

---

## Step 2: Convert Exam Scores to GPA

Raw exam scores in this dataset run from 0 to 100. To make them easier
to interpret, `scale_exam_score()` converts them to a GPA scale (0.0 to 4.0)
using the formula `(Exam_Score / 100) * 4`.

```{r gpa}
df <- scale_exam_score(df)
head(df[, c("Exam_Score", "gpa")])
```

This adds a new `gpa` column to the data frame which is used by several
other functions downstream.

---

## Step 3: Summarize Grades

`grade_analysis()` gives an overview of the GPA distribution across
all students.

```{r grades}
grade_analysis(df)
```

This gives you a quick sense of where most students fall on the GPA
scale and how much variation there is.

---

## Step 4: Explore Sleep Patterns

### Summary statistics

`sleep_summary()` returns the mean, median, and standard deviation of
student sleep hours across the whole dataset.

```{r sleep-summary}
sleep_summary(df)
```

### Relationship between sleep and study time

`sleep_study_relationship()` computes the Pearson correlation between
sleep hours and hours studied, and gives a plain-language interpretation.

```{r sleep-study}
sleep_study_relationship(df)
```

A correlation close to 0 here would suggest that how much a student
sleeps has little relationship with how much they study — students who
study more are not necessarily sleeping less.

---

## Step 5: Visualize Relationships

`plot_relationship()` lets you plot any two numeric variables against each other without writing any ggplot2 code yourself. You simply specify the variables you want to compare.

```{r plot-relationship}
plot_relationship(df, "Attendance", "gpa")
```

### Additional Visualization Functions

In addition to the flexible plotting function, the package also includes
predefined plots for common relationships.

Sleep vs GPA

```{r plot-sleep}
plot_sleep_vs_gpa(df)
```

This visualization highlights how sleep patterns may influence academic
outcomes, complementing the numerical summaries from earlier steps.

Together, these visualization functions allow users to both explore
custom relationships and quickly analyze key patterns built into the
package.

Study Time vs GPA

```{r plot-study}
plot_study_vs_gpa(df)
```

This plot focuses specifically on the relationship between study time
and GPA, making it easy to interpret one of the most important factors
in academic performance.


---

## Step 6: Predict GPA

`predict_gpa()` trains a linear regression model on the full dataset
and predicts GPA based on a new student's personal inputs. You just
provide your own values directly — no need to build a data frame.

```{r predict}
predict_gpa(hours_studied    = 20,
            sleep_hours      = 7,
            attendance       = 85,
            previous_scores  = 75,
            motivation_level = "Medium")
```

You can try different inputs to see how the prediction changes:

```{r predict-compare}
# High performing student
predict_gpa(hours_studied    = 35,
            sleep_hours      = 9,
            attendance       = 98,
            previous_scores  = 95,
            motivation_level = "High")

# Struggling student
predict_gpa(hours_studied    = 5,
            sleep_hours      = 4,
            attendance       = 60,
            previous_scores  = 45,
            motivation_level = "Low")
```

The model uses hours studied, sleep hours, attendance, previous scores,
and motivation level as predictors. It is a simple linear model - the
intention is to be able interpret the GPA rather than high predictive accuracy.

---

## Step 7: Calculate Your Performance Score

`performance_score()` combines key lifestyle variables into a single
score between 0 and 100. It also prints a performance message and
personalised suggestions based on your inputs.

Weights used:

- GPA: 30%
- Hours Studied: 25%
- Attendance: 15%
- Sleep Hours: 15%
- Extracurriculars: 15%

```{r performance-average}
# Average student
performance_score(
  gpa             = 2.5,
  hours_studied   = 15,
  attendance      = 75,
  sleep_hours     = 7,
  extracurricular = "No"
)
```

```{r performance-high}
# High achieving student
performance_score(
  gpa             = 3.8,
  hours_studied   = 35,
  attendance      = 95,
  sleep_hours     = 8,
  extracurricular = "Yes"
)
```

```{r performance-low}
# Struggling student
performance_score(
  gpa             = 1.5,
  hours_studied   = 5,
  attendance      = 60,
  sleep_hours     = 5,
  extracurricular = "No"
)
```

The function not only returns a score but also prints tailored suggestions
so users know exactly which habits to improve.

---
